what\_if\_1\_model\_datatable

### 2023-11-21 response_to_r1.qmd

---
title: Response to reviewers of "Spreadsheet modeling and wrangling with Python"
date: 2023-11-21
categories: [python, excel]
citation: false
output: html_document
format:
  html:
    embed-resources: true
---

@@

## Editor comments

**Comment 1**

I would like to apologize to the author for the delay in the review.  Although two reviewers accepted the assignment, one of them, at this point, is 65 days late with the review – therefore, I decided to proceed with a single review.

My first request to the author is to include other related ITE publications either as part of the literature review or to the other relevant sections.  Here is a short list of links to start:

- https://pubsonline.informs.org/doi/pdf/10.1287/ited.2021.0261
- https://pubsonline.informs.org/doi/full/10.1287/ited.2023.0285
- https://pubsonline.informs.org/doi/pdf/10.1287/ited.2022.0275

### 2024-02-23 python_excel_ite_final.tex
final revision

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Author style for INFORMS Transactions on Education (ited)
%% Mirko Janc, Ph.D., INFORMS, pubtech@informs.org
%% ver. 0.92, June 2009 -- default options: single-spaced, double-blinded
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\documentclass[ited]{informs3}                      % for a regular run
%\documentclass[ited,nonblindrev]{informs3}          % for review, not blinded
%\documentclass[ited,blindrev,copyedit]{informs3}    % spaced for copyediting
\documentclass[ited,blindrev]{informs3}              % for review, blinded

\usepackage[most]{tcolorbox}
\usepackage{graphicx}
%\usepackage{minted}
%\tcbuselibrary{minted} % tcolorbox minted library, required to use the "minted" tcb listing engine (this library is not loaded by the option [most])
\usepackage[colorlinks]{hyperref} % ALWAYS load this package LAST

@@

Numerous R and Python based data science courses have been developed and offered at universities and through online learning platforms such as \href{https://www.coursera.org/browse/data-science}{Coursera}, \href{https://www.edx.org/learn/data-science}{EdX}, \href{https://www.datacamp.com/}{DataCamp}, \href{https://www.udemy.com/topic/data-analysis/}{Udemy} and many others. One particularly relevant example are the courses in computation offered by the Naval Postgraduate School and described in \citet{alderson2022:interactive_computing}. In their courses, Jupyter notebooks (discussed in more detail in the next section) play a prominent role in providing an interactive development and learning environment for Python within an operations research context. 
In 2015, at Oakland University, we created a course called \href{http://www.sba.oakland.edu/faculty/isken/courses/pcda/}{Practical Computing for Data Analytics} (PCDA) that introduced both R and Python (along with a little Linux) for use in data analytics work. This new course was meant as a follow up to the \href{http://www.sba.oakland.edu/faculty/isken/courses/ba/}{spreadsheet based modeling course} we have had in place since 2001.

The PCDA course, like many introductory data science courses, focuses mostly on data wrangling, exploratory data analysis (EDA), and statistical/machine learning based predictive models. In teaching EDA using tools like \href{https://ggplot2.tidyverse.org/}{ggplot2}, students are challenged to think about how they would go about trying to use Excel to create some of the plots that are easy to do with R and Python based plotting libraries. There is some limited coverage of using Python to open Excel files and do worksheet manipulation. However, other than a Python based simulation of the Monte Hall 3-door problem, there was no real attempt to do the kind of modeling that is the focus of most spreadsheet based modeling textbooks.

@@

\begin{figure*}[!htbp]
\centering
\includegraphics[scale=0.5]{cells_edit_mode}
\caption{Notebook cells in edit mode}
\label{fig:cells_edit_mode}
\end{figure*}

\begin{figure*}[!htbp]
\centering
\includegraphics[scale=0.5]{cells_after_running}
\caption{Notebook cells after execution}
\label{fig:cells_after_running}
\end{figure*}

@@

\begin{figure}[!htbp]
	\includegraphics[width=1.0\textwidth]{objectives_whatif}
	\caption{Learning objectives for Submodule 1}
	\label{fig:objectives_whatif}
\end{figure}

@@

\begin{figure}[!htbp]
\includegraphics[width=1.0\textwidth]{output_2_0}
\caption{Typical spreadsheet model}
\label{fig:xl_model}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[scale=0.5]{model_inputs}
\caption{Instantiating model inputs and creating output formulas}
\label{fig:model_inputs}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[scale=0.5]{function_datatable}
\caption{A 2-way data table}
\label{fig:function_datatable}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics[scale=0.5]{datatable_scatter}
\caption{Scatter plot based on 2-way data table}
\label{fig:datatable_scatter}
\end{figure}

@@

For our initial design of a \code{BookstoreModel} class we make all of the base inputs class attributes (properties) and then add method attributes to compute outputs such as costs, revenues and profits. Students can easily see the mapping between elements in the initial procedural model and this new object oriented version. Several important concepts and syntactical details of doing object oriented programming Python are highlighted. Again, the goal is to weave in these more advanced Python programming concepts within the context of a very familiar modeling problem. This section of the notebook ends with using the \code{BookstoreModel} class to create a new model object instance with all of its base inputs instantiated with values. Before moving on, students are presented with a challenges involving enhancements to the \code{BookstoreModel} class. Interspersing short coding challenges throughout the notebook gives students a chance to test themselves on their understanding and to get some coding practice.

@@

This first notebook draws to a close with the creation of a \code{data\_table} function that allows us to do sensitivity analysis with the objected oriented model. The payoff is the creation of the following faceted plot showing how profit varies for different order quantities and demand levels. We recap the main things learned in this notebook and prepare to add goal seeking capability to our model.

\begin{figure}[!htbp]
\centering
\includegraphics[width=1.0\textwidth]{two_way_dt}
\caption{Faceted plot based on 2-way data table}
\label{fig:2way_dt}
\end{figure}

@@

A \code{goal\_seek} function is created that implements a simple bisection search and we use it to find the break even demand point in our bookstore model. With \code{data\_table} and \code{goal\_seek} functions implemented, we are ready to move on to doing Monte-Carlo simulation.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[scale=0.5]{oo_sim_design_specs}
\caption{Design specification for \code{simulate()} function}
\label{fig:oo_sim_design_specs}
\end{figure}

This is followed by a well commented first attempt at a \code{simulate} function. By this point, students are well equipped to understand the code as it builds on basic Python concepts and more advanced concepts that we have already covered in the EwP module. The rest of the notebook is spent trying out the \code{simulate()} function and ends with creating a grouped boxplot, see Figure \ref{fig:boxplot}, along with a faceted plot of profit histograms created with the popular \href{https://seaborn.pydata.org/}{Seaborn} package.

\begin{figure}[!htbp]
\centering
\includegraphics[width=1.0\textwidth]{boxplots}
\caption{Profit boxplots by order quantity}
\label{fig:boxplot}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[scale=0.5]{sim_wrapup}
\caption{Closing thoughts in simulation notebook}
\label{fig:sim_wrapup}
\end{figure}

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[scale=0.5]{objectives_packaging}
	\caption{Learning objectives for Submodule 2}
	\label{fig:objectives_packaging}
\end{figure}

@@

% Acknowledgments here
\ACKNOWLEDGMENT{%
% Enter the text of acknowledgments here
}% Leave this (end of acknowledgment)

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Author style for INFORMS Transactions on Education (ited)
%% Mirko Janc, Ph.D., INFORMS, pubtech@informs.org
%% ver. 0.92, June 2009 -- default options: single-spaced, double-blinded
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[ited]{informs3}                      % for a regular run
%\documentclass[ited,nonblindrev]{informs3}          % for review, not blinded
%\documentclass[ited,blindrev,copyedit]{informs3}    % spaced for copyediting
%\documentclass[ited,blindrev]{informs3}              % for review, blinded

\usepackage[most]{tcolorbox}
\usepackage{graphicx}
\usepackage{caption}
\newenvironment{codelisting}{\captionsetup{type=listing}}{}

\usepackage{minted}
%\tcbuselibrary{minted} % tcolorbox minted library, required to use the "minted" tcb listing engine (this library is not loaded by the option [most])

%\usepackage{listings}
%\usepackage{color}
%
%\definecolor{dkgreen}{rgb}{0,0.6,0}
%\definecolor{gray}{rgb}{0.5,0.5,0.5}
%\definecolor{mauve}{rgb}{0.58,0,0.82}
%
%\lstset{frame=tb,
%	language=Python,
%	aboveskip=3mm,
%	belowskip=3mm,
%	showstringspaces=false,
%	columns=flexible,
%	basicstyle={\small\ttfamily},
%	numbers=none,
%	numberstyle=\tiny\color{gray},
%	keywordstyle=\color{blue},
%	commentstyle=\color{dkgreen},
%	stringstyle=\color{mauve},
%	breaklines=true,
%	breakatwhitespace=true,
%	tabsize=3
%}

\usepackage[colorlinks]{hyperref} % ALWAYS load this package LAST

@@

Numerous R and Python based data science courses have been developed and offered at universities and through online learning platforms such as \href{https://www.coursera.org/browse/data-science}{Coursera}, \href{https://www.edx.org/learn/data-science}{EdX}, \href{https://www.datacamp.com/}{DataCamp}, \href{https://www.udemy.com/topic/data-analysis/}{Udemy} and many others. One particularly relevant example are the courses in computation offered by the Naval Postgraduate School and described in \citet{alderson2022:interactive_computing}. In their courses, Jupyter notebooks (discussed in more detail in the next section) play a prominent role in providing an interactive development and learning environment for Python within an operations research context. In \citet{pythonprojects:ite}, an analytics project based course is described in which Python plays a significant role. Students use Python for a number of data wrangling, exploratory data analysis and modeling tasks. Through a course like this, students gain valuable experience with a tool that is becoming more of an expectation in the analytics world. Another effective and fun way that students can gain experience with Python is through using it solve puzzles such as Wordle \citep{wordle_2023} or the Advent of Code \citep{AOC}.

In 2015, at Oakland University, we created a course called \href{http://www.sba.oakland.edu/faculty/isken/courses/pcda/}{Practical Computing for Data Analytics} (PCDA) that introduced both R and Python (along with a little Linux) for use in data analytics work. This new course was meant as a follow up to the \href{http://www.sba.oakland.edu/faculty/isken/courses/ba/}{spreadsheet based modeling course} we have had in place since 2001. The PCDA course, like many introductory data science courses, focuses mostly on data wrangling, exploratory data analysis (EDA), and statistical/machine learning based predictive models. In teaching EDA using tools like \href{https://ggplot2.tidyverse.org/}{ggplot2}, students are challenged to think about how they would go about trying to use Excel to create some of the plots that are easy to do with R and Python based plotting libraries. There is some limited coverage of using Python to open Excel files and do worksheet manipulation. However, other than a Python based simulation of the Monte Hall 3-door problem, there was no real attempt to do the kind of modeling that is the focus of most spreadsheet based modeling textbooks.

@@

\begin{figure*}[!htbp]
\centering
\includegraphics[scale=0.5]{images/cells_edit_mode}
\caption{Notebook cells in edit mode}
\label{fig:cells_edit_mode}
\end{figure*}

\begin{figure*}[!htbp]
\centering
\includegraphics[scale=0.5]{images/cells_after_running}
\caption{Notebook cells after execution}
\label{fig:cells_after_running}
\end{figure*}

@@

\begin{figure}[!htbp]
	\includegraphics[width=1.0\textwidth]{images/objectives_whatif}
	\caption{Learning objectives for Submodule 1}
	\label{fig:objectives_whatif}
\end{figure}

@@

\begin{figure}[!htbp]
\includegraphics[width=1.0\textwidth]{images/output_2_0}
\caption{Typical spreadsheet model}
\label{fig:xl_model}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[scale=0.5]{images/model_inputs}
\caption{Instantiating model inputs and creating output formulas}
\label{fig:model_inputs}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[scale=0.5]{images/function_datatable}
\caption{A 2-way data table}
\label{fig:function_datatable}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics[scale=0.5]{images/datatable_scatter}
\caption{Scatter plot based on 2-way data table}
\label{fig:datatable_scatter}
\end{figure}

@@

For our initial design of a \code{BookstoreModel} class we make all of the base inputs class attributes (properties) and then add method attributes to compute outputs such as costs, revenues and profits - see Figure \ref{fig:oo_model}. Students can easily see the mapping between elements in the initial procedural model and this new object oriented version. Several important concepts and syntactical details of doing object oriented programming Python are highlighted. Again, the goal is to weave in these more advanced Python programming concepts within the context of a very familiar modeling problem. This section of the notebook ends with using the \code{BookstoreModel} class to create a new model object instance with all of its base inputs instantiated with values. Before moving on, students are presented with a challenge involving enhancements to the \code{BookstoreModel} class. Interspersing short coding challenges throughout the notebook gives students a chance to test themselves on their understanding and to get some coding practice.

\begin{figure}[!htbp]
	\centering
	\includegraphics[scale=0.5]{images/oo_model}
	\caption{Object oriented version of the model}
	\label{fig:oo_model}
\end{figure}

@@

This first notebook draws to a close with the creation of a \code{data\_table} function (Figure \ref{fig:oo_datatable_function}) that allows us to do sensitivity analysis with the objected oriented model. The payoff is the creation of the faceted plot in Figure \ref{fig:2way_dt} showing how profit varies for different order quantities and demand levels. We recap the main things learned in this notebook and prepare to add goal seeking capability to our model.

\begin{figure}[!htbp]
	\centering
	\includegraphics[scale=0.5]{images/oo_datatable_function}
	\caption{An n-way DataTable function}
	\label{fig:oo_datatable_function}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics[width=1.0\textwidth]{images/two_way_dt}
\caption{Faceted plot based on 2-way data table}
\label{fig:2way_dt}
\end{figure}

@@

A \code{goal\_seek} function is created that implements a simple bisection search and we use it to find the break even demand point in our bookstore model. See Listing \ref{fig:goal_seek_function}. With \code{data\_table} and \code{goal\_seek} functions implemented, we are ready to move on to doing Monte-Carlo simulation.

\begin{listing}
\centering
\begin{minted}[fontsize=\tiny]{python}
def goal_seek(model, obj_fn, target, by_changing, a, b, N=100, verbose=False):
    '''Approximate solution of f(x)=0 on interval [a,b] by bisection method.

    Parameters
    ----------
    model : object
        User defined object containing the appropriate methods and properties for doing the desired goal seek
    obj_fn : function
        The function for which we are trying to approximate a solution f(x)=target.
    target : float
        The goal
    by_changing : string
        Name of the input variable in model
    a,b : numbers
        The interval in which to search for a solution. The function returns
        None if (f(a) - target) * (f(b) - target) >= 0 since a solution is not guaranteed.
    N : (positive) integer
        The number of iterations to implement.
    verbose : boolean (default=False)
        If True, root finding progress is reported

    Returns
    -------
    x_N : number
        The midpoint of the Nth interval computed by the bisection method. 
        If all signs of values f(a_n), f(b_n) and f(m_n) are the same at any
        iteration, the bisection method fails and return None.
    '''  
    # Clone the model
    model_clone = copy.deepcopy(model)

    # The following bisection search is a direct adaptation of
    # https://www.math.ubc.ca/~pwalls/math-python/roots-optimization/bisection/  
    setattr(model_clone, by_changing, a)
    f_a_0 = getattr(model_clone, obj_fn)()
    setattr(model_clone, by_changing, b)
    f_b_0 = getattr(model_clone, obj_fn)()

    if (f_a_0 - target) * (f_b_0 - target) >= 0:
        # print("Bisection method fails.")
        return None

    # Initialize the end points
    a_n = a
    b_n = b
    for n in range(1, N+1):
        # Compute the midpoint
        m_n = (a_n + b_n)/2

        # Function value at midpoint
        setattr(model_clone, by_changing, m_n)
        f_m_n = getattr(model_clone, obj_fn)()

        # Function value at a_n
        setattr(model_clone, by_changing, a_n)
        f_a_n = getattr(model_clone, obj_fn)()

        # Function value at b_n
        setattr(model_clone, by_changing, b_n)
        f_b_n = getattr(model_clone, obj_fn)()

        if verbose:
            print(f"n = {n}, a_n = {a_n}, b_n = {b_n}, m_n = {m_n}, width = {b_n - a_n}")

        # Figure out which half the root is in, or if we hit it exactly, or if the search failed
        if (f_a_n - target) * (f_m_n - target) < 0:
            a_n = a_n
            b_n = m_n
        elif (f_b_n - target) * (f_m_n - target) < 0:
            a_n = m_n
            b_n = b_n
        elif f_m_n == target:
            return m_n
        else:
            return None

    # If we get here we hit iteration limit, return best solution found so far
    return (a_n + b_n)/2
\end{minted}
\caption{\code{goal\_seek} function}
\label{fig:goal_seek_function}
\end{listing}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[scale=0.5]{images/oo_sim_design_specs}
\caption{Design specification for \code{simulate()} function}
\label{fig:oo_sim_design_specs}
\end{figure}

This is followed by a well commented first attempt at a \code{simulate} function. By this point, students are well equipped to understand the code as it builds on basic Python concepts and more advanced concepts that we have already covered in the EwP module. See Listing \ref{fig:simulate_function}.

\begin{listing}
\centering
\begin{minted}[fontsize=\tiny]{python}
def simulate(model, random_inputs, outputs, scenario_inputs=None, keep_random_inputs=False):
    '''Simulate model for one or more scenarios

    Parameters
    ----------
    model : object
        User defined object containing the appropriate methods and properties for computing outputs from inputs
    random_intputs : dict of str to sequence of random variates
        Keys are stochastic input variable names and values are sequence of $n$ random variates, where $n$ is the number of simulation replications
    outputs : list of str
        List of output variable names
    scenario_inputs : optional (default is None), dict of str to sequence
        Keys are deterministic input variable names and values are sequence of values for each scenario for this variable. Is consumed by
        scikit-learn ParameterGrid() function. See https://scikit-learn.org/stable/modules/generated/sklearn.model_selection.ParameterGrid.html
    keep_random_inputs : optional (default is False), boolean
        If True, all the random input variates are included in the results dataframe

    Returns
    -------
    results_df : list of dict
        Values of all outputs for each simulation replication. If `scenario_inputs` is not None, then this is also for every combination of scenario inputs
    '''

    # Clone the model
    model_clone = copy.deepcopy(model)

    # Update clone with random_inputs
    model_clone.update(random_inputs)

    # Store raw simulation input values if desired
    if keep_random_inputs:
        scenario_base_vals = vars(model_clone)
    else:
        scenario_base_vals = vars(model)

    # Initialize output counters and containers
    scenario_num = 0
    scenario_results = []

    # Check if multiple scenarios
    if scenario_inputs is not None:
        # Create parameter grid for scenario inputs
        sim_param_grid = list(ParameterGrid(scenario_inputs))

        # Scenario loop
        for params in sim_param_grid:
            model_clone.update(params)
            # Initialize scenario related outputs
            result = {}
            scenario_vals = copy.copy(params)
            result['scenario_base_vals'] = scenario_base_vals
            result['scenario_num'] = scenario_num
            result['scenario_vals'] = scenario_vals
            raw_output = {}

            # Output measure loop
            for output_name in outputs:
                output_array = getattr(model_clone, output_name)()
                raw_output[output_name] = output_array

            # Gather results for this scenario
            result['output'] = raw_output
            scenario_results.append(result)
            scenario_num += 1

        return scenario_results

    else:
        # Similar logic to above, but only a single scenario
        results = []
        result = {}

        result['scenario_base_vals'] = scenario_base_vals
        result['scenario_num'] = scenario_num
        result['scenario_vals'] = {}

        raw_output = {}
        for output_name in outputs:
            output_array = getattr(model_clone, output_name)()
            raw_output[output_name] = output_array

        result['output'] = raw_output  
        results.append(result)

        return results
\end{minted}
\caption{\code{simulate} function}
\label{fig:simulate_function}
\end{listing}

The rest of the notebook is spent trying out the \code{simulate()} function and ends with creating a grouped boxplot, see Figure \ref{fig:boxplot}, along with a faceted plot of profit histograms created with the popular \href{https://seaborn.pydata.org/}{Seaborn} package.

\begin{figure}[!htbp]
\centering
\includegraphics[width=1.0\textwidth]{images/boxplots}
\caption{Profit boxplots by order quantity}
\label{fig:boxplot}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[scale=0.5]{images/sim_wrapup}
\caption{Closing thoughts in simulation notebook}
\label{fig:sim_wrapup}
\end{figure}

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[scale=0.5]{images/objectives_packaging}
	\caption{Learning objectives for Submodule 2}
	\label{fig:objectives_packaging}
\end{figure}

@@

% Acknowledgments here
%\ACKNOWLEDGMENT{%
% Enter the text of acknowledgments here
%}% Leave this (end of acknowledgment)
